Estimating duration using survival analysis

### 2024-03-22 cis-perfect-testing.tex
Fix refs
Ch 5 Overleaf comments

When estimating incidence from prevalence, the mean duration, not its median, is the most important feature of the duration distribution~\autocite{freemanPrevalence}.
Duration distributions tend to be right-skewed.
Therefore, estimates of the mean are sensitive to the tail of the duration distribution.

In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
However, this study suffered from a small sample size and only 20 days of follow-up, enabling estimation of the tail.
Therefore, model assumptions drive the size of the estimated tail.

In this chapter, I develop a method to estimate the duration distribution using CIS data.
The CIS sampling is less frequent, but with a large sample size and long follow-up.

I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (\cref{perf-test:sec:problem}).
I use a simulation study (\cref{perf-test:sec:simulation-study}) to evaluate a range of priors for the survival distribution (\cref{perf-test:sec:parameters-priors}).
These include a prior to combine the information in the CIS with the results of \cref{E-ATACCC}.
Reasonable priors successfully recover the true distribution.
However, the method must be able to account for false negatives, discussed in \cref{E-imperf-test}.

@@

The CIS tests each individual infrequently, either weekly or monthly (see \cref{E-intro:sec:cis}).
Infrequent testing complicates analysis of the data in two ways.
First, infection episodes can go undetected.
Second, the infection episode start and end times are known imprecisely.
This section explains these issues. The rest of the chapter develops a method to tackle  them.

\subsection{Undetected infections}

The first complication in the analysis is that infection episodes can be undetected.
Undetected infection episodes are episodes with no associated positive tests.
Therefore, there is no information directly observed about the episode; even the number that occur is unobserved.
They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
Undetected episodes are important because they are shorter than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration had been 24 days or more, the individual would still be detectable at their next test and detected then.
CIS has many undetected infections.
This is because the interval between tests in the same individual (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

\subsection{Unobserved episode beginning and end times} \label{perf-test:sec:interval-censoring}

The second complication is that the exact beginning and end times of detected infection episodes are unknown.
The imprecision in the beginning of the infection episode arises because the only observation that episode $j$ has started is that an individual was negative at one test, on day $l_j^{(b)}-1$ (the minus one is because the first day that the infection could have begun is the day after the negative test), and positive at the next test, on day $r_j^{(b)}$.
Similarly, the end of the infection episode is bounded by the final positive test on day $l_j^{(e)}$ and the following negative test on day $r_j^{(e)}+1$ (the plus one is because the last day that the infection could have ended is the day before the negative test).
\Cref{perf-test:fig:double-interval-censor} shows this issue graphically.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

\subsection{Approach} \label{perf-test:sec:approach}

I base my analysis on the 4,800 episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
That is, the individual tested negative before and after the episode.
Without a negative before the episode there is no lower bound on the episode's start time.
It could be arbitrarily long and there is little information on its length.
A negative test follows each episode unless the individual was lost to follow-up before their episode ended.
This is rare.
Episodes removed in this step are incorporated in the adjustment for undetected episodes.

\section{Survival analysis}

In survival analysis, the event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
It is used in many areas, including how long a machine lasts before breaking down or how long a patient spends in hospital after a treatment.
Here, I am estimating the distribution of the duration of positivity.
The initiating event is the episode's start; the terminating event is the episode's end.

Undetected events where there is no information on these events is known as \emph{truncation}.
A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
There is a subtlety that makes the CIS different from standard left truncation.
In CIS, individuals in which undetected infections occurred are enrolled in the study.
Therefore, some data has been collected about them.
In standard left truncation, nothing is known about the missed events.
Here, I will incorporate the additional information into the analysis.
I refer to these events as ``undetected'', rather than truncated, to emphasise this difference.

The time of the initiating and/or terminating events are not known exactly.
This is known as \emph{censored} in the survival analysis literature.
Censored events are present in almost all survival analyses.
An event is \emph{left censored} if it is known to occur before a certain time.
An event is \emph{right censored} if it is known to occur after a certain time.
An event is \emph{interval censored} if it is known to occur within an interval.
%Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
As explained in \cref{perf-test:sec:problem}, the initiating and terminating events are only bounded, and hence interval censored.
The data is \emph{doubly censored} because both the initiating and terminating events are censored.
To handle doubly censored data, the times need to be jointly modelled~\autocite[and references therein]{liSemiparametric}.
In CIS data both the initiating and the terminating events are interval censored.
Therefore, the CIS data is \emph{double interval censored}.
\Textcite{sunAnalysis,bogaertsSurvival} review methods for double interval censored data.

Few studies look at the combination of both double interval censoring and undetected (or truncated) events.
This is especially true within the human biostatistical literature.
The literature does include theoretical frameworks which include the double censored and truncated case~\autocite{turnbullEmpirical,dempsterMaximum}, yet these frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
They condition on the truncation (a selection effect) and the period where the initiating event occurs.
They condition on the initiating event time because it provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
This is not the case for the CIS.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.

Censored and truncated data are normal for estimation of the survival time of bird and insect nests~\autocite{heiseyABCs}.
\textcite{heiseyModelling} developed a framework for handling this situation.
The framework is general, allowing for arbitrary censoring patterns and double interval censoring.
They apply it to a situation where the visit times (possible detection times) are global, simplifying the likelihood.
I adapt this framework for CIS, when each individual has their own visit schedule (see section \cref{perf-test:sec:model}).

I use Bayesian inference for two reasons.
First, frequentist estimators are poorly developed and their properties (\eg consistency and convergence rate) are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
Second, a prior can include information from previous studies, such as from \cref{E-ATACCC}.
Alternatively, the prior can favour a smoothly changing hazard.
I explore priors in section 1.4.
These prior structures overcome issues caused by the undetected episodes ~\autocite{caoBias}.

Bayesian methods have been used for nest survival before~\autocite{heBayesiana,heBayesian,caoModeling} by augmenting the data with the unobserved times of infection.
The augmentation creates a parameter per detected infection.
This increases the dimensionality of the problem greatly, making it computationally infeasible within the SRS.
I developed a novel Bayesian framework without explicit data augmentation.
This method scales to the large number of episodes present in CIS.

@@

In this section, I develop a model accounting for double interval censoring and undetected episodes.
This approach extends the framework of \textcite{heiseyModelling} to include individuals without detected episodes.
This framework is itself based on long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}.
I augment the data set with the full number of episodes, including those that are undetected.

Integrating over the number of missed episodes is possible analytically.
This makes inference tractable within modern, high-performance Bayesian software such as Stan.
Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
Let $\Ncis$ be the number of individuals in the cohort and $\nsched$ be the number of test schedules.
%Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
Let $i = 1, \dots, \nsched$ index infection schedules and $j = 1, \dots \ntot$ index infection episodes.

All the relevant information about episode $i$ can be fully characterized by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode being and end respectively and $\sched_j$ is the set of test times for the individual in which the episode occurs.
The episode's duration is $d_j = e_j - b_j + 1$.
%, the number of days the individual in which the episode occurred, was positive for 
This triplet is assumed iid.

$b_j$ and $e_j$ span possible infection and recovery times.
$\sched_j$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.
Over the short period of time considered, each unique test schedule has zero or one detected episodes.

Let $i \in \set{D}$ indicate $\sched_i$ is associated with a detected episode.
That is, $i \in \set{D}$ if and only if there exists a $j(i)$ such that $b_{j(i)}$ and $e_{j(i)}$ such that $(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode.
$(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode when the following three conditions are met.
First, $\exists t \in \sched_i$ such that $b_{j(i)} \leq t \leq e_{j(i)}$; this condition is equivalent to having at least one positive test for the episode.
Second, $\exists t \in \sched_i$ such that $t < b_{j(i)}$, to lower bound the start of the episode.
Finally, $\exists t \in \sched_i$ such that $t > e_{j(i)}$, to upper bound the end of the episode; this condition will be met for any episode sufficiently far in the past in individuals not lost to follow-up (almost always the case in this context).
For a new context, including recent infections, this condition can be relaxed by considering these episodes as right censored, at little cost except a more complicated likelihood expression.

To relate the state space to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), for a test schedule $\sched_i$.
Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the model parameters, $\theta$.

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
  This space is empty for $i \notin \set{D}$; in this case, $n_{ia} = 0$ admissible episodes occurred in individuals with test schedule $\sched_i$.
  Otherwise, $i \in \set{D}$ then $n_{ia} =1$ admissible episode occurred in an individual with test schedule $\sched_{i}$.
  Conditional on the episode occurring in individuals with test schedule $\sched_i$, it is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
  \label{perf-test:def:admissible}
\item
  Undetected episodes, $\Omega_i^C$, which have an infection and recovery time such that they would not have tested positive.
  An unknown number, $n_{iu}$ of these observed in individuals with test schedule $\sched_i$.
  Conditional on the episode occurring in individuals with test schedule $\sched_i$, it is undetected with probability $p_{iu} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
\item
  Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
  detected.
  These are all remaining episodes not in the previous sets.
  $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
  Conditional on the episode occurring in these individuals, it is inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \theta)$.
\end{itemize}

% The detected region, $\Omega_i = \alpha_i \cup \beta_i$, for test schedule $\sched_i$ is the region where we could have observed episodes. 
% The notation is for consistency with \textcite{heiseyModelling}.
% $\Omega_i$ is the complement of $\Omega^C_i$.

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
\caption[Admissible, inadmissible, and undetected infections]{The regions (as defined in the main text) for an individual with test schedule $\sched_i$ whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. Inadmissible regions are unshaded. The black (impossible) region is because it does not make sense for recovery to occur prior to infection. \label{perf-test:fig:partitionSpace}}
\end{figure}

@@

Any episode $j$ independently belongs to one of these classes.
I assume the individual in which each episode occurs is iid uniform across the cohort.
Therefore, conditional on $\ntot$, the number of episodes in each class, $(n_{ia}, n_{iu}, n_{i\bar{a}})^T$, are distributed multinomially.
The above probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$) are conditional on the episode occurring with test schedule $\sched_i$.
I multiply these probabilities by $\eta_i/\Ncis$, where $\eta_i$ is the number of individuals with test schedule $\sched_i$, to give the probability that an episode belongs to each class, without conditioning on a specific individual.

The undetected classes for each individual cannot be distinguished by any observable data.
Each class's count needs to be marginalised over to calculate the likelihood.
Because the counts have a joint multinomial distribution, this is equivalent to only considering the sum of the counts.
The overall probability of an episode being unobserved is $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\nsched} \eta_i p_{iu}$.
Furthermore, the inadmissible classes are guaranteed to have zero episodes.
Counts of zero do not contribute to a multinomial pmf (they multiply by 1).
Therefore, I combine the undetected classes into a single class and ignore the inadmissible classes.

Let $\na = (n_{1a}, \dots, n_{\nsched a})^T$ be the vector of counts of episodes in the admissible classes.
The counts $\na$ and $\nnodet$ are jointly distributed multinomially.
The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{\nsched} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
Hence:
\begin{align}
p(\na \mid \ntot, \theta)
&= p(\nnodet, \na\mid \ntot, \theta) \\
&= \frac{\ntot!}{\nnodet! \prod_{i=1}^{\nsched} n_{ia}!}  \pnodet^{\nnodet}\prod_{i=1}^{\nsched} \left(\frac{1}{\nsched}p_{ia}\right)^{n_{ia}} \\
&\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \label{perf-test:eq:augmented-likelihood}
\end{align}
the final line follows because $\nnodet = \ntot - \ndet$, $n_{ia} = 1$ for $i \in \set{D}$, and $n_{ia} = 0$ otherwise.

@@

I implemented the inference in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
Convergence was assessed using Rhat and ESS as described in \cref{E-inc-prev:sec:MCMC}, and all runs checked for divergent transitions.

>>>>

When estimating incidence from prevalence, the mean duration, not its median, is the most important feature of the duration distribution~\autocite{freemanPrevalence}.
Duration distributions tend to be right-skewed.
Therefore, estimates of the mean would be sensitive to the tail of the duration distribution.

In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
However, this study suffered from a small sample size and only 20 days of follow-up, preventing accurate estimation of the tail.
Therefore, model assumptions drive the size of the estimated tail.

In this chapter, I develop a method to estimate the duration distribution using CIS data.
The CIS sampling is less frequent, but with has a large sample size (\numprint{4800} detected infection episodes) and a long follow-up.

In \cref{perf-test:sec:problem}, I describe the challenges analysing the data presents.
This motivates using a survival analysis framework which I introduce in \cref{perf-test:sec:surv-analysis}, before deriving the statistical model for the CIS setting in \cref{perf-test:sec:model}.
Next, in \cref{perf-test:sec:parameters-priors}, I consider how to parameterise the survival function and choose priors; these include vague and smoothing priors as well as including information from \cref{E-ATACCC}.
I evaluate the model and priors using a simulation study (\cref{perf-test:sec:simulation-study}), showing that they perform well.
Finally, I discuss the results in \cref{perf-test:sec:discussion}.

@@

The CIS tests each individual infrequently, either weekly or monthly (see \cref{E-intro:sec:cis}), complicating the data in two ways.
First, infection episodes can go undetected.
Second, the infection episode start and end times are known imprecisely.
I explain these issues below, and devote the rest of the chapter to the development of a method to tackle  them.

\subsection{Undetected infections} \label{perf-test:sec:undetected}

The first complication in the analysis is that infection episodes can be undetected.
Undetected infection episodes are episodes with no associated positive tests.
Therefore, there is no information directly observed about the episode; even the number that occur is latent.
They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
Undetected episodes are important because they are, on average, shorter than detected episodes; therefore, neglecting them would lead to a biased estimate of the survival distribution.
To see why this is the case, consider the infection episode in figure \cref{perf-test:fig:truncation}.
Here two individuals are shown, with identical data collected for each of them (a series of negative
tests).
The fist individual was never infected whereas the second individual was; the latter is an undetected infection. 
The undetected infection episode begins at time 33 and has a 10-day duration.
The individual would have been detected at their next test if their duration had instead been 24 days or more.
The gap between tests in the CIS (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
Therefore, it is likely that many infection in the CIS cohort were undetected.
This is because the interval between tests in the same individual (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{Graphical illustration of an undetected infection episode, represented by the shaded area. \label{perf-test:fig:truncation}}
\end{figure}

\subsection{Unobserved times for each episode's beginning and end} \label{perf-test:sec:interval-censoring}

The second complication is that the beginning and end times of detected infection episodes is latent.
Recall, from \cref{E-biology-data:sec:cis-episodes}, that the time episode $j$ starts, $b_j$, is only known to lie in the interval $[l_j^{(b)}, r_j^{(b)}]$ where $l_j^{(b)}$ is the day after the negative test prior to the start of episode $j$ and $r_j^{(b)}$ is the first positive test in episode $j$.
Similarly, for the end of the infection episode, $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
\Cref{perf-test:fig:double-interval-censor} shows this issue graphically.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=\textwidth]{cis-perfect-testing/double-interval-censor}}
  \caption[Double-interval censoring in CIS data]{%
        The beginning and end of each episode observed in the CIS data is only known up to an interval.
        Demonstrated here by a participant who is recorded negative at time 7 and positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode.
    }
    \label{perf-test:fig:double-interval-censor}
\end{figure}

\section{Survival analysis} \label{perf-test:sec:surv-analysis}

In survival analysis, the interest is in estimating the distribution of the interval between an \emph{initiating event} and a \emph{terminating event}.
In this context, the initiating event is the episode's start; the terminating event is the episode's end.
I start by introducing the relevant, standard concepts in survival analysis; a wide range of texts cover them including \textcite[chapter 1]{bogaertsSurvival}.

A common issue is that the initiating and/or terminating events are not known exactly.
This is known as the event being \emph{censored}.
An event is \emph{left censored} if it is known to occur before a certain time.
An event is \emph{right censored} if it is known to occur after a certain time.
An event is \emph{interval censored} if it is known to occur within an interval.
If both the initiating and terminating event are censored, the data are \emph{doubly censored}.
To handle doubly censored data, the times need to be jointly modelled~\autocite[and references therein]{liSemiparametric}.
If both events are interval censored then the data are \emph{doubly interval censored}.

Sometimes, only intervals satisfying some property are recorded.
In this case, data are \emph{truncated}.
\emph{Left truncation} is when only events lasting longer than a minimum amount of time are observed.
\emph{Right truncation} is when only events lasting shorter than a maximum amount of time are observed.
Without correcting for the truncation, inference will be biased.

In the CIS data, the initiating and terminating events are only bounded (see \cref{perf-test:sec:interval-censoring}).
Therefore, the data are doubly interval censored.

There is also undetected infections in the CIS (see \cref{perf-test:sec:undetected}).
However, the situation is subtly different from the standard truncation situation encountered in the literature.
In the standard situation, nothing is known about the unrecorded intervals.
But in CIS, individuals in which undetected infections occurred are enrolled in the study.
Therefore, we know the times that they were tested.
This provides information on how long their infections may have been.
I refer to these events as ``undetected'', rather than truncated, to emphasise this difference.

\Textcite{sunAnalysis,bogaertsSurvival} review methods for doubly interval censored data.
However, few studies look at data both doubly interval censored and with undetected (or truncated) events.
% This is especially true within the biostatistical literature on human diseases.
General theoretical frameworks, which include this case, have been developed~\autocite{turnbullEmpirical,dempsterMaximum}.
Yet only the special case where the terminating event is either uncensored or right censored has been applied in practice~\autocite[e.g.][]{sunEmpirical,bacchettiNonparametric,shenNonparametric}.
These studies base their inference on a likelihood contribution for each detected infection episode conditional on $b_j \in [l_j^{(b)}, r_j^{(b)}]$.

For this conditional likelihood to be valid, they assume this conditioning provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
This assumption performs well in these settings, when the length of the intervals is fairly constant.
In particular, when the interval lengths are near-constant, $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) \approx p(\text{episode $j$ detected})$.
However, the interval lengths change from one to four weeks in the CIS (see \cref{E-intro:sec:cis}).

Infection episodes in an individual's one week testing period are over-represented in the detected infections.
This is because more frequent testing will detect a higher proportion of infections.
Therefore, for the average detected episode, $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) > p(\text{episode $j$ detected})$.
Therefore, conditioning on $b_j \in [l_j^{(b)}, r_j^{(b)}]$ is inappropriate.

Doubly interval censored and truncated data are common when assessing the survival time of bird and insect nests~\autocite{heiseyABCs}.
\textcite{heiseyModelling}, building on on long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}, develops a likelihood-based inference framework, allowing for arbitrary truncation patterns and double interval censoring.
They apply it to a situation where the visit times (possible detection times) are common to all the individuals, simplifying the likelihood.
I adapt this framework for CIS, when each individual has their own visit schedule (see section \cref{perf-test:sec:model}).

I use Bayesian inference for two reasons.
First, the frequentist properties (\eg consistency and convergence rate) of proposed estimators are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
Second, a prior can include information from previous studies, such as from \cref{E-ATACCC}.
Alternatively, the prior can favour a smoothly changing hazard, which is biologically sensible, and can overcome issues caused by the undetected episodes ~\autocite{caoBias}.

Bayesian methods have been used for nest survival before~\autocite{heBayesiana,heBayesian,caoModeling} by augmenting the data with the unobserved times of infection.
The data augmentation approach creates a parameter for each detected infection.
This increases the dimensionality of the problem greatly, making it computationally infeasible with the large sample size of the CIS within the SRS.

@@

In this section, I develop a statistical model accounting for double interval censoring and undetected episodes.
This approach extends the framework of \textcite{heiseyModelling} to include all individuals in the CIS, including those without detected episodes.

The number of missed episodes is now a latent quantity that needs to be considered.
Marginalising over the number of missed episodes is possible analytically.
This makes inference tractable using the large sample size and the computing resources available within the SRS (see \cref{E-biology-data:sec:SRS}).
In particular, the posterior can be expressed within modern, high-performance Bayesian software such as Stan.

The model will be used to analyse on the 4,800 episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
A negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
Similar information comes from the negative test after the end of the episode.
This is typically observed, unless the individual is lost to follow-up before their episode is ended, which happens rarely.
I will incorporate episodes excluded in this step in the model for undetected infections.

\subsection{Types of episodes}

Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
Let $\Ncis$ be the number of individuals in the cohort and $\nsched$ be the number of distinct test schedules.
A test schedule, $\sched_i$, is the set of days an individual in the cohort is tested at.
Each individual has exactly one test schedule but a test schedule could have multiple individuals associated with it.
A test schedule, $\sched_i$ is the set of all times 
%Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
Let $i = 1, \dots, \nsched$ index infection schedules and $j = 1, \dots \ntot$ index infection episodes.

All the relevant information about the infection episode $j$ can be fully characterized by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode begin and end respectively and $\sched_j$ is the set of test times for the individual in which the episode occurs.
The episode's duration is $d_j = e_j - b_j + 1$ (see sec 3.?).
%, the number of days the individual in which the episode occurred, was positive for 
This triplet is assumed iid.

The times $b_j$ and $e_j$ span possible infection and recovery times.
$\sched_j$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.

Over the short calendar period considered, each individual has zero or one detected episodes.
Let $i \in \set{D}$ indicate $\sched_i$ is associated with a detected episode.
That is, $i \in \set{D}$ if and only if there exists a $j(i)$ such that $b_{j(i)}$ and $e_{j(i)}$ such that $(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode. 
This happens
$(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode when all the following conditions are met.
\begin{enumerate}
    \item $\exists t \in \sched_i$ such that $b_{j(i)} \leq t \leq e_{j(i)}$; this condition is equivalent to having at least one positive test for the episode.
    \item $\exists t \in \sched_i$ such that $t < b_{j(i)}$, lower bounding the start of the episode.
    \item $\exists t \in \sched_i$ such that $t > e_{j(i)}$, upper bounding the end of the episode.
\end{enumerate}
For a new context, including recent infections, this condition can be relaxed by considering these episodes as right censored, at little cost except a more complicated likelihood expression.

To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_j$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$).
Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\theta$, discussed in ??.

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
  This space is empty for $i \notin \set{D}$; in this case, $n_{ia} = 0$ admissible episodes occurred in individuals with test schedule $\sched_i$.
  Otherwise, if $i \in \set{D}$ then $n_{ia} =1$ admissible episode occurred in an individual with test schedule $\sched_{i}$.
  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
  \label{perf-test:def:admissible}
\item
  Undetected episodes, $\Omega_i$, which have an infection and recovery time such that they would not have tested positive.
  An unknown number, $n_{iu}$ of these occur in individuals with test schedule $\sched_i$.
  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is undetected with probability $p_{iu} = \prob((b, e) \in \Omega_i \mid \theta)$.
\item
  Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
  detected.
  These are all remaining episodes not in the previous sets, with $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \theta)$.
\end{itemize}

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
\caption[Admissible, inadmissible, and undetected infections]{The regions (as defined in the main text) for an individual with test schedule $\sched_i$ whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. Inadmissible regions are unshaded. The black (impossible) region covers the cases in which recovery occurs prior to infection. \label{perf-test:fig:partitionSpace}}
\end{figure}

@@

Any episode $j$ independently belongs to one of these classes.
I assume the individual in which each episode occurs is iid uniform across the cohort.
Therefore, conditional on $\ntot$, the number of episodes in each class, $(n_{ia}, n_{iu}, n_{i\bar{a}})^T$, are distributed multinomially.
with probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$), which are conditional on these episode occurring with test schedule $\sched_i$.
By multiplying these probabilities by $\eta_i/\Ncis$, where $\eta_i$ is the number of individuals with test schedule $\sched_i$, the probability that an episode belongs to each class, without conditioning on a specific individual can be derived.

{\color{red} Would this be the place to introduce a new subsection? on how to calculate the likelihood? Or on line 242?}
The undetected classes for each individual cannot be distinguished by any observable data.
Each class's count needs to be marginalised over to calculate the likelihood.
Because the counts have a joint multinomial distribution, this is equivalent to only considering the sum of the counts.
The overall probability of an episode being unobserved is $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\nsched} \eta_i p_{iu}$.
Furthermore, the inadmissible classes are guaranteed to have zero episodes.
Counts of zero do not contribute to the multinomial probability mass function (they multiply by 1).
Therefore, I combine the undetected classes into a single class and ignore the inadmissible classes.

{\color{red} subsection here instead of line 233?}
Let $\na = (n_{1a}, \dots, n_{\nsched a})^T$ be the vector of counts of episodes in the admissible classes.
The counts $\na$ and $\nnodet$ are jointly distributed multinomially.
The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{\nsched} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
Hence:
\begin{align}
p(\na \mid \ntot, \theta)
&= p(\nnodet, \na\mid \ntot, \theta) \\
&= \frac{\ntot!}{\nnodet! \prod_{i=1}^{\nsched} n_{ia}!}  \pnodet^{\nnodet}\prod_{i=1}^{\nsched} \left(\frac{1}{\nsched}p_{ia}\right)^{n_{ia}} \\
&\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \label{perf-test:eq:augmented-likelihood}
\end{align}
the final line follows because $\nnodet = \ntot - \ndet$, $n_{ia} = 1$ for $i \in \set{D}$, and $n_{ia} = 0$ otherwise.

@@

I implemented the inference in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
Convergence was assessed using Rhat and ESS as described in \cref{E-MCMC:sec:convergence}, and all runs checked for divergent transitions.
